Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-10 paper/paper.tex
Stuff

\noindent
It is important to highlight that, in our mechanisation, these proofs follow
\emph{exactly} the steps that one would do in a pen-and-paper proof.
We show the series of rewrite steps for \coq{hylo_fusion_l} in Figure~\ref{fig:proof-fusion-l}:
\begin{figure}
\begin{coqcode}
(* E2 : f2 \o g1 =e g2 \o fmap f2 *)
(* ------------------------------------- *)
(* Goal : f2 \o hylo g1 h1 =e hylo g2 h1 *)

@@

Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
  (* [knapsack_alg wvs] as a respecful morphism *)
\end{coqcode}
The hylomorphism for knapsack is as follows, where \coq{out_nat} is the
recursive coalgebra for \coq{nat}.
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
Coq's code extraction mechanism is unable to inline several definitions in this
case. We have manually inlined the extracted code for simplicity. The reader
can check in our artefact that the extracted code can be trivially
inlined to produce the following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) =
    (let rec f x0 =
      if x0=0 then Uint63.of_int (0)
      else let fn := f (x0-1) in { lFix_out = { 
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.

@@

\mvol{Do we need to discuss this or can we cut it?}

In particular, the problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.

>>>>

\noindent
It is important to highlight that, in our mechanisation, these proofs follow
\emph{exactly} the steps that one would do in a pen-and-paper proof.  We show
the series of rewrite steps for \coq{hylo_fusion_l} in
Figure~\ref{fig:proof-fusion-l}. The steps of this proof are: (1) apply the
uniqueness law of recursive hylomorphisms; (2) rewrite the goal using the
property that functors preserve composition; (3) rewrite the goal using the
condition of \coq{hylo_fusion_l}; (4) use the uniqueness law of hylomorphisms to
fold $a \circ F \; f \circ c$ into $f$.
\begin{figure}
\begin{coqcode}
(* E2 : f2 \o g1 =e g2 \o fmap f2 *)
(* ------------------------------------- *)
(* Goal : f2 \o hylo g1 h1 =e hylo g2 h1 *)

@@

Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
  (* [knapsack_alg wvs] as a respecful morphism *)
\end{coqcode}
The hylomorphism for knapsack is as follows, where \coq{out_nat} is the
recursive coalgebra for \coq{nat}.
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
Coq's code extraction mechanism is unable to inline several definitions in this
case. We have manually inlined the extracted code for simplicity. The reader
can check in our artefact that the extracted code can be trivially
inlined to produce the following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) =
    (let rec f x0 =
      if x0=0 then
        { lFix_out = {
             shape = Uint63.of_int (0);
             cont  = fun _ -> f 0 } }
      else let fn := f (x0-1) in { lFix_out = { 
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.

@@

In particular, the problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.
